\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-27 paper.tex
Clear up a few TODOs

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. Once the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example genome ARG (gARG) embedded in a pedigree.
(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \noderef{a} to \noderef{p}. Black lines show inheritance paths connecting
genomes in the current generation (\noderef{a} to \noderef{d}) with their ancestors.
Genomes \noderef{a} and \noderef{c} are the product of two independent
recombination events (red) between
the paternal genomes \noderef{e}
and \noderef{f}, and regions of genome inherited are shown with shaded colour.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

@@

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% TODO more?
mathematically defined~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep{song2004minimum,song2005efficient,rasmussen2014genome,heine2018bridging,ignatieva2021kwarg}.
In this Griffiths encoding, as well as nodes marking sampling events,
we have two additional node types in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG (eARG). (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
for a discussion of this important point).
}
\end{figure}

@@

% \cite{shipilina2023origin} discuss the idea of a ``haplotype block''
% (equivalent to the ``bricked tree sequence'' of
% \cite{nowbandegani2023extremely},
% % TODO look this up. I expect it's basically the same??
% and the block-based ARG data structure of
% \citep{palamara2016argon}),
% here we consider the unique sets of
% samples that coalesce at a node in the ARG over a particular
% genome interval, and the fundamental limitations they place
% on ARG inference.
% [FIXME find somewhere natural for this. The last clause is too vague - what
% are the implications?]
% Alternative gARG encodings split transmitted intervals and associate each with
% a new node, resulting in a single genome being represented by a group of
% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for

%%%% Some old text about the coalescent we might still want to plunder
% % If we want to use the coalescent, then really the assumptions
% % need to make sense.
% % n << Ne is a silly assumption in today's datasets
% A core assumption of the coalescent is that the sample size $n$
% is much less than the effective population size, $N_e$.
% Several human datasets now consist of hundreds of thousands of
% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
% and so sample size is substantially \emph{larger} than $N_e$
% (often assumed to be $10^4$ in humans).
% Agricultural datasets are even more extreme.
% For example,
% the US dairy cattle database alone currently comprises more than 6 million
% animals with SNP array
% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
% while the effective population size in modern dairy cattle breeds is
% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
% An extreme sign of these breeding practices is that there are only two ancestral
% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
% comprises close to 7000 genomes, which are part of multi-generation pedigrees
% with millions of animals and extensive SNP array genotype and phenotype
% data \citep[e.g.][]{Cesarani2022}.
% In another example, genome and SNP array data for 440,610 individuals within
% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
% Effective population size in modern pig breeds is also less than 100 due to
% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
% Another core assumption of the coalescent model is that the genome (or
% at least the region under study) is short enough that the number of extant
% lineages remains much smaller than $N_e$ at all times. Whole
% genome sequences have been available for model organisms
% for over a decade now, % True? Citation?
% and indeed complete chromosome-level assemblies are possible
% in humans~\citep{miga2020telomere}.
% Projects are under way to obtain high-quality assemblies
% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
% and ultimately worldwide~\citep{lewin2022earth}.

@@

% TODO revisit number here
\section*{Appendix}
\appendix

>>>>

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
An example genome ARG (gARG) embedded in a pedigree.
(A) Diploid individuals (blue), visualised in a highly inbred pedigree and
labelled $D_1$ to $D_8$,
contain both paternal and maternal  genomes
labelled \noderef{a} to \noderef{p}. Black lines show inheritance paths connecting
genomes in the current generation (\noderef{a} to \noderef{d}) with their ancestors.
Genomes \noderef{a} and \noderef{c} are the product of two independent
recombination events (red) between
the paternal genomes \noderef{e}
and \noderef{f}, and regions of genome inherited are shown with shaded colour.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
(C) The corresponding local trees.
}
\end{figure}

@@

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% TODO more?
mathematically defined~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
In this Griffiths encoding, as well as nodes marking sampling events,
we have two additional node types in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG (eARG). (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
for a discussion of this important point).
}
\end{figure}

@@

@@

\section*{Appendix}
\appendix
